---
title: "Penguin Cleaning"
output: html_document
date: '2023-01-30'
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```
## Starting off

In R there is "base R" which is what you get when you
you use the library() function to load additional
packages that you install.  

One way to make sure that your work is reproducible is
to keep the details well organized and to record
exactly what you do and why.  One way to help with this
is as you go on and, potentially, need to add new libraries
use the second code chunk to keep track of 
all the libraries that you use.  This makes it 
easier for other people who want to reproduce your
code -- and for yourself when you come back to the
code months later. 

Use the second chunk for this because you can
use the first chunk to set any options you want
for your whole document. 

The code chunk below shows how to load palmerpenguins.
Also load dplyr, tidyr, skimr, ggplot2, stringr.

```{r}
library(palmerpenguins)

```

You can learn more about the palmerpenguins
data by looking at this page https://allisonhorst.github.io/palmerpenguins.
The penguin package contains two data sets:
`penguins` and `penguins_raw`.  Use the View() 
function to look at the two data sets. You can also
read over the help for the two data sets (`?penguins` `?penguins_raw`  )

Our goal is to go from *penguins_raw *to a new 
data set *penguins_new* that is as close as you can get it 
to identical to penguins.   We can test if they
are the same using the code `identical(penguins, penguins_new)`.
You can also use `all.equal(penguins, penguins_new)`. 

## First step

Look closely at both data frames.  Make a list of things you will need to do to the raw data set to make it match the cleaned data. 
Make the list as detailed as possible.  You may come up with more items later but get started before you do anything else.  Also you
don't need to worry about the order of the steps.  And write in English, not code.

## My list of tasks

1. 
2.
3.
4.
5. 


## Comparing data frames

You can "eye ball" the two data frames to see some differences.  However, you can also use code to do the comparisons.
Using code will generally help you spot differences that are not obvious, but are important.

This blog post examines the comparison of data frames in much more detail.
https://sharla.party/post/comparing-two-dfs/ (Worth reading and playing
with.)

Towards the bottom of this document there is a code chunk called "test".
This contains the test code mentioned above.
That way you can periodically run it as you are working.   

###  What happens if you run that code chunk now?



In the code chunk below create penguins_new from
the penguins_raw data set using the assignment operator `<-`. 

```{r}

```
Now run your test code chunk again.

In the code chunk below use base functions to get a sense of the data.  (See slides)

### What are some things you notice?


```{r}

```

Another way to get a sense of a data set is to use
the `skim()` function from the `skimr` package.
Remember: How do you access a package? 

Do that below.

```{r}

```

### Do you notice anything new? What?



## Start cleaning

Now, working on cleaning the data.  Let's work both independently and collaboratively. 

### Useful functions
Here are some functions you may want to use, you can learn more about these by using the help system. 
Typing `?tolower` in the console would be an example of that.

     Useful functions:
     tolower(x) 
      whatever you put inside gets changed to lower case (there is also toupper to go the other way)
     
     substr(x, start, stop)  
     extract the substring of x staring with the “start” position and going to “stop” so substr(x, 1,3)  where x is “elephant” would extract “elep”.
     
     sub(pattern, replacement, x) 
      lets you replace a specific string with another one. Usually we will want to add fixed = TRUE   
     sub(pattern, replacement, x, fixed = TRUE) 
     sub(“el”, “bell”, x, fixed = TRUE)  would give you bellephant.   
     
     gsub() lets you do the same as sub but with multiple instances
     
     as.factor(x)  is used to convert a variable to a factor.
     
     as.numeric(x) is used to convert a variable to numeric.
     
     as.integer(x) is used to convert a variable to an integer.
     
     relocate() is used to rearrange the order of the columns
     
     names()  names(df) gives the column names of a data frame (or vector) in the form of a vector.
     names(df) <- vector_of_names
     Lets us assign names to the columns of a data frame (or vector)
     
     typeof(df$(pick your column) is used when you want to check what data type is your column. For example is your column numeric, or a string type etc.(df  is where you include your dataset) 
     
     str_to_lower/upper()  is used to convert a string to lower case or upper case ( similar functions are ‘tolower()/to upper)
     
     Getting year from a date object
     
     format(x, "%Y") 
     #  (note these are “curly quotes” and won’t work in R, you need to use "straight quotes".  Don’t copy and paste.  Also %Y is part of date formatting and means 4 digit year.  For two digit year you would use %y.  Same for month and day (%M, %m, %D, %d)
     
     Good website on working with strings in R
     
     https://github.com/Cghlewis/data-wrangling-functions/wiki/Strings

### dplyr

For cleaning data the tidyverse package dplyr is extremely useful.  It makes R work similarly to 
SQL.

Here are the main "dplyr verbs." (This nice
summary is copied from Alison Horst's dplyr tutorial.) 

- `filter()`: keep rows that satisfy your conditions
- `select()`: keep or exclude some columns
- `rename()`: rename columns
- `relocate()`: move columns around
- `mutate()`: add a new column
- `group_by()` + `summarize()`: get summary statistics by group
- `across()`: apply a function across columns
- `count()`: quickly find counts for different groups
- `case_when()`: like friendly if-else

We'll also want to use various other functions for specific tasks.

## Getting to work!

Working below in well organized code chunks modify  your penguins_new file.  (I have started you with
a few examples.)

Sometimes you may make a mistake. But by carefully tracking what you do you can always go back
and run all of the code, starting at the top and get back to the point where you need to fix 
things. 

Also, dplyr is commonly used with the pipe operator.
This is really useful, but we while we are working it is good to avoid chains that are too long. Long
pipe chains are hard to debug.

A piping example. Notice that penguins_new in the environment doesn't change. 

### What is this code doing? Can you add more to it to continue your work?

```{r eval=FALSE}
penguins_new |> 
     select(-studyName, -`Sample Number`)  
```

To make it update the penguins_new object, assign
the code back to the object. After running
the next chunk, check the data in the environment.


```{r}
penguins_new <- penguins_new |> 
                select(-studyName, -`Sample Number`) 
```
Update the above code chunk to get rid of any other
columns you don't need.  But be careful.

Now let's try another pretty simple one, rename().

Change the code with some examples. When you 
have it the way you want, assign your changes
back to the object.

```{r}
penguins_new |> rename(newname = oldname)
```

Don't worry about doing all of them at this point because you can always go back and make changes.

The last example I'll include is mutate, which you can use to create new variables or modify
existing variables.

Let's say we want to create year, what do we
need to do? How can we find a function to do 
what we want to do?

```{r}
penguins_raw |>  
     mutate(year = WHAT, sex = WHAT)
```

Now continue working on your own for a while. 
The important thing is to work iteratively, always  checking your partial results.


One concept you might need at the end is `attributes`.  You can get the listing of attributes for a data frame (or any object in R)
with `attributes(df)`.  

You can remove, modify or add an attribute with the `attr()` function.

setting a value (overwrites anything that is there already).

attr(df , "attribute name") <- "newvalue"

Deleting an attribute completely.
attr(df , "attribute name") <- NULL


### Test code chunk.

```{r test}
identical(penguins, penguins_new)
all.equal(penguins, penguins_new)


```



You will finish this up on your own and push to github. I'll help with that 
next week, please come with your file ready to push if you haven't done that on your own.

As a special challenge, now add a new variable
to your data set based on data from the raw data that
is not currently included. 





